Determinism, Complexity, and Predictability in Computer Performance

### 2014-08-04 wpe.tex
pushing changes midstream, in case of lossage
Liz final read through round I

Estimating the entropy of an arbitrary, real-valued time series is a
challenging problem.  Our approach to this problem, which is the topic
of this section, draws upon methods and results from a variety of
fields including time-series analysis, dynamical systems, and
stochastic processes.

@@

From a different point of view, we can consider the information (as
measured by the Shannon entropy) contained in a single observable, say
the \emph{present}, of the system. This information can be partitioned
into two components: the information shared with past
observations---e.g., the mutual information between the past and
present---and the information in the present that is not contained in
the past (aka ``the conditional entropy of the present given the
past'').  The first part is known as the \emph{redundancy}, as it is
information in the present that is also in the past.  The second part
is the aforementioned Shannon entropy rate.  It seems obvious that the
more redundancy in a signal, the more predictable it should be.
Redundancy is critical to prediction, in general.  And the specific
\emph{form} of the redundancy should dictate whether a particular
prediction method will work well or poorly on the corresponding
signal.  A linear method cannot capture or make use of nonlinear
redundancy, for instance.  This issue---which has been explored
extensively using clean data and/or systems where the generating
process is known, but not so much with empirical data from unknown
systems---is central to the claims in this paper and the discussion in
the following section.

@@

Here we will be utilizing a variation of the basic permutation entropy
technique, the \emph{weighted permutation entropy} (WPE), which was
introduced in~\cite{fadlallah2013}.  The intent behind the weighting
is to correct for observational noise that is larger than the trends
in the data, but smaller than the larger-scale features.  Consider,
for example, a signal that switches between two fixed points and
contains some additive noise. The PE of such a signal will be
dominated by the noise about the fixed points, driving it to $\approx
1$---which in some sense hides the fact that the signal is actually
quite structured.  In this situation, the terms of the weighted
permutation entropy are dominated by the switching rather than the
stochastic fluctuations. To accomplish this, the \emph{weight} of a
permutation is taken into account:
\begin{align*}
  w(x_{i+1}^\ell) = \frac{1}{\ell} \sum_{j = i + 1}^{i+\ell}
                      \left( x_j - \bar{x}_{i+1}^\ell \right)^2
\end{align*}
where $x_{i+1}^\ell$ is a sequence of values $x_{i+1}, \ldots,
x_{i+\ell}$, and $\bar{x}_{i+1}^\ell$ is the arithmetic mean of
those values.

@@

In practice, calculating permutation entropy and weighted permutation
entropy involves choosing a good value for the word length $\ell$. The
primary consideration in that choice is that the value be large enough
that forbidden ordinals are discovered, yet small enough that
reasonable statistics over the ordinals are gathered.  If an average
of 100 counts per ordinal is considered to be sufficient, for
instance, then $\ell = \operatornamewithlimits{argmax}_{\hat{\ell}} \{
N \gtrapprox 100 \hat{\ell}! \}$.  In the literature, $3 \le \ell \le
6$ is a standard choice---generally without any formal justification.
In theory, the permutation entropy should reach an asymptote with
increasing $\ell$, but that can require an arbitrarily long time
series. In practice, what one should do is calculate the
\emph{persistent} permutation entropy by increasing $\ell$ until the
result converges, but data length issues can intrude before that
convergence is reached.  We used this approach, which represents a
good balance between accurate ordinal statistics and finite-data
effects, to choose $\ell$ values for the experiments in the following
section.

>>>>

Estimating the entropy of an arbitrary, real-valued time series is a
significant challenge.  Our approach to this problem, which is the topic of
this section, draws upon methods and results from a variety of fields
including time-series analysis, dynamical systems, and stochastic
processes.

@@

From a different point of view, we can consider the information (as
measured by the Shannon entropy) contained in a single observable, say
the \emph{present}, of the system. This information can be partitioned
into two components: the information shared with past
observations---e.g., the mutual information between the past and
present---and the information in the present that is not contained in
the past (aka ``the conditional entropy of the present given the
past'').  The first part is known as the \emph{redundancy}:
information in the present that is also in the past.  The second part
is the aforementioned Shannon entropy rate.  It seems obvious that the
more redundancy in a signal, the more predictable it should be.  And
the specific \emph{form} of the redundancy should dictate whether a
particular prediction method will work well or poorly on the
corresponding signal.  A linear method cannot capture or make use of
nonlinear redundancy, for instance.  This issue---which has been
explored extensively using clean data and/or systems where the
generating process is known, but not so much with empirical data from
unknown systems---is central to the claims in this paper and the
discussion in the following section.

@@

Here we will be utilizing a variation of the basic permutation entropy
technique, the \emph{weighted permutation entropy} (WPE), which was
introduced in~\cite{fadlallah2013}.  \alert{We never actually say WHY
  we use it.}  The intent behind the weighting is to correct for
observational noise that is larger than the trends in the data, but
smaller than the larger-scale features.  Consider, for example, a
signal that switches between two fixed points and contains some
additive noise. The PE of such a signal will be dominated by the noise
about the fixed points, driving it to $\approx 1$---which in some
sense hides the fact that the signal is actually quite structured.  In
this situation, the terms of the weighted permutation entropy are
dominated by the switching rather than the stochastic fluctuations. To
accomplish this, the \emph{weight} of a permutation is taken into
account:
\begin{align*}
  w(x_{i+1}^\ell) = \frac{1}{\ell} \sum_{j = i + 1}^{i+\ell}
                      \left( x_j - \bar{x}_{i+1}^\ell \right)^2
\end{align*}
where $x_{i+1}^\ell$ is a sequence of values $x_{i+1}, \ldots,
x_{i+\ell}$, and $\bar{x}_{i+1}^\ell$ is the arithmetic mean of
those values.

@@

In practice, calculating permutation entropy and weighted permutation
entropy involves choosing a good value for the word length $\ell$. The
primary consideration in that choice is that the value be large enough
that forbidden ordinals are discovered, yet small enough that
reasonable statistics over the ordinals are gathered.  If an average
of 100 counts per ordinal is considered to be sufficient, for
instance, then $\ell = \operatornamewithlimits{argmax}_{\hat{\ell}} \{
N \gtrapprox 100 \hat{\ell}! \}$.  In the literature, $3 \le \ell \le
6$ is a standard choice---generally without any formal justification.
In theory, the permutation entropy should reach an asymptote with
increasing $\ell$, but that can require an arbitrarily long time
series. In practice, what one should do is calculate the
\emph{persistent} permutation entropy by increasing $\ell$ until the
result converges, but data length issues can intrude before that
convergence is reached.  We used this approach, which we believe
strikes a good balance between accurate ordinal statistics and
finite-data effects, to choose $\ell$ values for the experiments in
the following section.
